Syntax and Typing for Core Cedille

### 2018-03-26 docs/cedille-core/spec.tex
about to move the spec for Cedille core to github, so pushing some old edits
git-svn-id: https://svn.divms.uiowa.edu/repos/clc/projects/lambda-encodings/implementations/cedille@29940 bd178f5f-7bea-4014-844f-495ca19b8805

\begin{figure}
\[
\begin{array}{llll}
  \textit{term variables}\ u & \ & \ &\ \\
  \textit{type variables}\ X & \ & \ &\ \\
  \textit{kind variables}\ k & \ & \ &\ \\
  \textit{variables}\ x & ::= & u\ |\ X\ |\ k &\ \\  
  \textit{pure terms}\ p & ::= & u\ |\ p\ p'\ |\ \absu{\lambda}{u}{p} &\ \\
  \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a term variable} \\
  \ &\ & \star&\textnormal{kind for types}\\
  \ &\ & \Box&\textnormal{sole superkind} \\
  \ &\ & t.1 & \textnormal{project first view of a dependent intersection}\\
  \ &\ & t.2 & \textnormal{project second view of a dependent intersection}\\
  \ &\ & \beta\{t\} &\textnormal{proof of a true equation, where the proof erases to }t  \\
  \ &\ & \varsigma\ t &\textnormal{symmetry of equality}  \\
  \ &\ & t\ t' & \textnormal{application of term to term}\\
  \ &\ & t\ \mhyph t' & \textnormal{application of a term to an erased argument}\\
  \ &\ & \rho\ t\ @\ x.t'\ \mhyph\ t''&\textnormal{equality elimination by type-guided rewriting}  \\ 
  \ &\ & \abs{\forall}{x}{t}{t'}\ &\textnormal{implicit product (quantify over erased argument)} \\
  \ &\ & \abs{\Pi}{x}{t}{t'}\ &\textnormal{explicit product (usual }\Pi\textnormal{-type)} \\
  \ &\ & \abs{\iota}{x}{T}{T'}\ &\textnormal{dependent intersection} \\
  \ &\ & \abs{\lambda}{x}{t}{t'} & \textnormal{usual }\lambda\textnormal{-abstraction} \\
  \ &\ & \abs{\Lambda}{x}{t}{t'} & \textnormal{erased }\lambda\textnormal{-abstraction}\\
  \ &\ & [ t , t' \ @\ x.t'' ] & \textnormal{introduce dependent intersection}\\
  \ &\ & \phi\ t\ \mhyph\ t'\ \{t''\} & \textnormal{when }t\textnormal{ proves }t'\textnormal{ and }t''
                                   \textnormal{ are equal, erase to }t'' \\
  \ & \ & [ x = t : t' ]\ \mhyph\ t'' & \textnormal{let }x\textnormal{ equal }t\textnormal{ of type }t'\textnormal{ in }t'' \\
  \ &\ & \{ p \simeq p' \} &\textnormal{equality between pure terms}
\end{array}
  \]
  \caption{Sytax for core Cedille}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{lll}
    |x| & = & x \\
    |\star| & = & \star \\
    |\Box| & = & \Box \\
    |t.1| & = & |t| \\
    |t.2| & = & |t| \\
    |\beta\{t\}| & = & |t|\\
    |\varsigma\ t| & = & |t| \\
    |t\ t'| & = & |t|\ |t'| \\
    |t\ \mhyph t'| & = & |t| \\
    |\rho\ t\ @\ x.t'\ \mhyph\ t''| & = & |t''| \\
    |\abs{\forall}{x}{t}{t'}| & = & \abs{\forall}{x}{|t|}{|t'|}\\
    |\abs{\Pi}{x}{t}{t'}| & = & \abs{\Pi}{x}{|t|}{|t'|}\\
    |\abs{\iota}{x}{T}{T'}| & = & \abs{\iota}{x}{|t|}{|t'|} \\
    |\abs{\lambda}{x}{t}{t'}| & = &  \absu{\lambda}{x}{|t'|} \\
    |\abs{\Lambda}{x}{t}{t'}| & = &  |t'| \\ 
    |[ t , t' \ @\ x.t'' ]| & = & |t| \\ 
    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| \\
    |[ x = t : t' ]\ \mhyph\ t'' | & = & (\absu{\lambda}{x}{|t''|})\ |t|\\
    |\{ t \simeq t' \}|| & = & \{ |t| \simeq |t'| \}
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{lll}
    \infer{\Gamma\vdash \star : \Box}{\ } &
    \infer{\Gamma\vdash\abs{\Pi}{x}{t}{t'} : s'}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : s' \quad \textit{Var}(x,s)} &
    \infer{\Gamma\vdash\abs{\forall}{x}{t}{t'} : \star}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : \star \quad \textit{Var}(x,s)} 
    \\ \\
    \infer{\Gamma, x : t \vdash x : t}{\ } &
    \infer{\Gamma\vdash t\ t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\Pi}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash\abs{\iota}{u}{t}{t'} : \star}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : \star} 
    \\ \\
    \infer{\Gamma\vdash t.2 : [t/x]t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
    \infer{\Gamma\vdash\abs{\lambda}{x}{t}{t'} : \abs{\Pi}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\Pi}{x}{t}{t''} : s}  &
    \infer{\Gamma\vdash\abs{\Lambda}{x}{t}{t'} : \abs{\forall}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\forall}{x}{t}{t''} : s} \\ \\
    \infer{\Gamma\vdash t.1 : t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &
    \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2 \quad \Gamma \vdash \abs{\iota}{x}{t_1}{t_2} : \star}
    \\ \\
    \infer{\Gamma\vdash \beta\{t\} : \{ t' \simeq t'\}}{\Gamma\vdash \{ t' \simeq t' \} : \star } &
    \infer{\Gamma\vdash \{ p \simeq p'\} : \star}{\textit{FV}(p\ p')\subseteq \textit{dom}(\Gamma)} &
    \infer{\Gamma\vdash \rho\ t\ @\ x.t'\ \mhyph\ t'' : [t_2/x]t'}{\Gamma \vdash t : \{ t_1 \simeq t_2 \} \quad \Gamma\vdash t'' : [t_1/x]t'}
    \\ \\
    \infer{\Gamma\vdash \varsigma\ t : \{ t_2 \simeq t_1\}}{\Gamma\vdash t : \{t_1 \simeq t_2\}} &
    \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'} &
    \infer{\Gamma\vdash [ k = t_1 : \Box ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : \Box \quad
                                                               \Gamma, k = t_1 : t' \vdash t_2 : t''} 
    \\ \\
    \multicolumn{2}{l}{
    \infer{\Gamma\vdash [ x = t_1 : t' ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma \vdash t' : s \quad \textit{Var}(x,s) \quad
      \Gamma, x = t_1 : t' \vdash t_2 : t''} } &
    \     
  \end{array}
  \]
  \caption{Type-checking rules for Core Cedille}
  \label{fig:rules}
\end{figure}

@@

  \end{itemize}
\end{itemize}

>>>>

\begin{figure}
\[
\begin{array}{llll}
  \textit{term variables}\ u & \ & \ &\ \\
  \textit{type variables}\ X & \ & \ &\ \\
  \textit{kind variables}\ k & \ & \ &\ \\
  \textit{variables}\ x & ::= & u\ |\ X\ |\ k &\ \\  
  \textit{pure terms}\ p & ::= & u\ |\ p\ p'\ |\ \absu{\lambda}{u}{p} &\ \\
  \textit{annotated terms}\ t & ::= & x\ & \textnormal{use of a variable} \\
  \ &\ & \star&\textnormal{kind for types}\\
  \ &\ & \Box&\textnormal{sole superkind} \\
  \ &\ & t.1 & \textnormal{project first view of a dependent intersection}\\
  \ &\ & t.2 & \textnormal{project second view of a dependent intersection}\\
  \ &\ & \beta\ t'\{t\} &\textnormal{proof of }t' \simeq t'\textnormal{, where the proof erases to }t  \\
  \ &\ & \varsigma\ t &\textnormal{symmetry of equality}  \\
  \ &\ & t\ t' & \textnormal{application of term to term}\\
  \ &\ & t\ \mhyph t' & \textnormal{application of a term to an erased argument}\\
  \ &\ & \rho\ t\ @\ x.t'\ \mhyph\ t''&\textnormal{equality elimination by type-guided rewriting}  \\ 
  \ &\ & \abs{\forall}{x}{t}{t'}\ &\textnormal{implicit product (quantify over erased argument)} \\
  \ &\ & \abs{\Pi}{x}{t}{t'}\ &\textnormal{explicit product (usual }\Pi\textnormal{-type)} \\
  \ &\ & \abs{\iota}{x}{T}{T'}\ &\textnormal{dependent intersection} \\
  \ &\ & \abs{\lambda}{x}{t}{t'} & \textnormal{usual }\lambda\textnormal{-abstraction} \\
  \ &\ & \abs{\Lambda}{x}{t}{t'} & \textnormal{erased }\lambda\textnormal{-abstraction}\\
  \ &\ & [ t , t' \ @\ x.t'' ] & \textnormal{introduce dependent intersection}\\
  \ &\ & \phi\ t\ \mhyph\ t'\ \{t''\} & \textnormal{when }t\textnormal{ proves }t'\textnormal{ and }t''
                                   \textnormal{ are equal, erase to }t'' \\
  \ & \ & [ x = t : t' ]\ \mhyph\ t'' & \textnormal{let }x\textnormal{ equal }t\textnormal{ of type }t'\textnormal{ in }t'' \\
  \ &\ & \{ p \simeq p' \} &\textnormal{equality between pure terms}
\end{array}
  \]
  \caption{Sytax for core Cedille}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{lll}
    |x| & = & x \\
    |\star| & = & \star \\
    |\Box| & = & \Box \\
    |t.1| & = & |t| \\
    |t.2| & = & |t| \\
    |\beta\ t'\{t\}| & = & |t|\\
    |\varsigma\ t| & = & |t| \\
    |t\ t'| & = & |t|\ |t'| \\
    |t\ \mhyph t'| & = & |t| \\
    |\rho\ t\ @\ x.t'\ \mhyph\ t''| & = & |t''| \\
    |\abs{\forall}{x}{t}{t'}| & = & \abs{\forall}{x}{|t|}{|t'|}\\
    |\abs{\Pi}{x}{t}{t'}| & = & \abs{\Pi}{x}{|t|}{|t'|}\\
    |\abs{\iota}{x}{T}{T'}| & = & \abs{\iota}{x}{|t|}{|t'|} \\
    |\abs{\lambda}{u}{t}{t'}| & = &  \absu{\lambda}{u}{|t'|} \\
    |\abs{\lambda}{X}{t}{t'}| & = &  \abs{\lambda}{X}{|t|}{|t'|} \\
    |\abs{\Lambda}{x}{t}{t'}| & = &  |t'| \\ 
    |[ t , t' \ @\ x.t'' ]| & = & |t| \\ 
    |\phi\ t\ \mhyph\ t'\ \{t''\}| & = & |t''| \\
    |[ x = t : t' ]\ \mhyph\ t'' | & = & (\absu{\lambda}{x}{|t''|})\ |t|\\
    |\{ t \simeq t' \}|| & = & \{ |t| \simeq |t'| \}
  \end{array}
  \]
  \caption{Erasure for annotated terms}
  \label{fig:eraser}
\end{figure}

@@

\begin{figure}
  \[
  \begin{array}{lll}
    \infer{\Gamma\vdash \star : \Box}{\ } &
    \infer{\Gamma\vdash\abs{\Pi}{x}{t}{t'} : s'}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : s' \quad \textit{Var}(x,s)} &
    \infer{\Gamma\vdash\abs{\forall}{x}{t}{t'} : \star}{\Gamma \vdash t : s \quad \Gamma,x:t\vdash t' : \star \quad \textit{Var}(x,s)} 
    \\ \\
    \infer{\Gamma, x : t \vdash x : t}{\ } &
    \infer{\Gamma\vdash t\ t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\Pi}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash\abs{\iota}{u}{t}{t'} : \star}{\Gamma \vdash t : \star \quad \Gamma,x:t\vdash t' : \star} 
    \\ \\
    \infer{\Gamma\vdash t.2 : [t/x]t_2}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &    
    \infer{\Gamma\vdash\abs{\lambda}{x}{t}{t'} : \abs{\Pi}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad \Gamma \vdash \abs{\Pi}{x}{t}{t''} : s}  &
    \infer{\Gamma\vdash \varsigma\ t : \{ t_2 \simeq t_1\}}{\Gamma\vdash t : \{t_1 \simeq t_2\}}

\\ \\
    \infer{\Gamma\vdash t.1 : t_1}{\Gamma\vdash t : \abs{\iota}{x}{t_1}{t_2}} &
    \infer{\Gamma\vdash t\ \mhyph t' : [t'/x]t_2}{\Gamma\vdash t : \abs{\forall}{x}{t_1}{t_2} & \Gamma\vdash t : t_1} &
    \infer{\Gamma\vdash \rho\ t\ @\ x.t'\ \mhyph\ t'' : [t_2/x]t'}{\Gamma \vdash t : \{ t_1 \simeq t_2 \} \quad \Gamma\vdash t'' : [t_1/x]t'}
    \\ \\
    \infer{\Gamma\vdash \beta\ t'\{t\} : \{ t' \simeq t'\}}{\Gamma\vdash \{ t' \simeq t' \} : \star } &
    \infer{\Gamma\vdash \{ p \simeq p'\} : \star}{\textit{FV}(p\ p')\subseteq \textit{dom}(\Gamma)} &
     \infer{\Gamma\vdash [ k = t_1 : \Box ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : \Box \quad
                                                               \Gamma, k = t_1 : t' \vdash t_2 : t''}  
    \\ \\
    \multicolumn{2}{l}{
    \infer{\Gamma\vdash [t,t'\ @\ x.t_2] : \abs{\iota}{x}{t_1}{t_2}}{\Gamma\vdash t : t_1 \quad \Gamma\vdash t' : [t/x]t_2 \quad \Gamma \vdash \abs{\iota}{x}{t_1}{t_2} : \star \quad |t| = |t'|}} &  
  \infer{\Gamma\vdash \phi\ t\ \mhyph\ t_1\ \{t_2\} : t'}{\Gamma \vdash t : \{t_1 \simeq t_2\} \quad \Gamma\vdash t_1 : t'}  
    \\ \\

    \multicolumn{2}{l}{
    \infer{\Gamma\vdash [ x = t_1 ]\ \mhyph\ t_2 : t''}{\Gamma\vdash t_1 : t' \quad \Gamma \vdash t' : s \quad \textit{Var}(x,s) \quad
      \Gamma, x = t_1 : t' \vdash t_2 : t''} } &
   \ 
    \\ \\
    \multicolumn{3}{l}{
      \infer{\Gamma\vdash\abs{\Lambda}{x}{t}{t'} : \abs{\forall}{x}{t}{t''}}{\Gamma,x:t\vdash t':t'' \quad x\not\in\textit{FV}(|t'|)\quad \Gamma \vdash \abs{\forall}{x}{t}{t''} : s}}
  \end{array}
  \]
  \caption{Type-checking rules for Core Cedille}
  \label{fig:rules}
\end{figure}

@@

  \end{itemize}
\item Feb. 14, 2018, pm:
  \begin{itemize}
  \item Corrected glaring omission of the requirement for typing $[t,t'\ @\ x.T]$, that $|t| = |t'|$.  It is fine if one implements
    this check instead as $|t| =_{\beta\eta} |t'|$, simply to save writing a separate function for testing $\alpha$-equivalence. 
  \end{itemize}
\item Feb. 16:
  \begin{itemize}
  \item removed type annotation on definiens in a let-expression.
  \item Changed syntax $\beta\{t\}$ to $\beta\ t'\{t\}$.
    \end{itemize}
\item Feb. 20: fix silly typo noticed by Victor, in the rule for $t.2$.  Also, erase $\lambda x$ differently from $\lambda X$.

\item Feb. 21: fixed rule for $\Lambda$-terms: we must check that the bound variable is not free in the erasure of the body.
\end{itemize}
